---
title: "Cardiogenic shock Call Data"
output: html_notebook
---

```{r}
# for data manipulation and visualization
library(tidyverse)
# for flexible paths and reproducible research
library(here)
library(tableone)
library(dplyr)
library(knitr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(stringr)

```

```{r}
 # Load data
Total_Data <- read_csv(here("data", "scdata.csv"), show_col_types = FALSE)
Total_Data <- Total_Data %>% 
  filter(if_any(everything(), ~ !is.na(.)))

```

## Data basics
population stats of sex, age, diagnosis, and CS Stage
```{r}
# Load the data


# Total participants
n_total <- nrow(Total_Data)

# Mean ± SD for age
age_stats <- Total_Data %>%
  summarise(
    Mean = round(mean(age, na.rm = TRUE), 1),
    SD = round(sd(age, na.rm = TRUE), 1)
  ) %>%
  mutate(Label = "Age (Mean ± SD)",
         Value = paste0(Mean, " ± ", SD)) %>%
  select(Label, Value)

# Sex breakdown
sex_counts <- Total_Data %>%
  count(Sex) %>%
  mutate(Percent = round(100 * n / sum(n), 1),
         Label = paste("Sex:", Sex),
         Value = paste0(n, " (", Percent, "%)")) %>%
  select(Label, Value)

# Diagnosis breakdown
diagnosis_counts <- Total_Data %>%
  count(Diagnosis) %>%
  mutate(Label = paste("Diagnosis:", Diagnosis),
         Percent = round(100 * n / sum(n), 1),
         Value = paste0(n, " (", Percent, "%)")) %>%
  select(Label, Value)

# CS Stage breakdown (excluding NAs)
cs_stage_counts <- Total_Data %>%
  filter(!is.na(`CS Stage`)) %>%
  count(`CS Stage`) %>%
  mutate(Label = paste("CS Stage:", `CS Stage`),
         Percent = round(100 * n / sum(n), 1),
         Value = paste0(n, " (", Percent, "%)")) %>%
  select(Label, Value)

# Combine all into one summary table
summary_table <- bind_rows(
  tibble(Label = "Number of Participants", Value = paste0("n = ", n_total)),
  age_stats,
  sex_counts,
  diagnosis_counts,
  cs_stage_counts
)

# Display
kable(summary_table, col.names = c("Variable", "Value"), caption = "Population Summary Table")

```


## CS Stage by CS, MCS, and Other
This is the graph representation 
```{r}
# Step 1: Prepare summarized data with standard error (placeholder using sqrt(n))
stage_summary <- Total_Data %>%
  filter(!is.na(`CS Stage`), !is.na(Diagnosis)) %>%
  count(Diagnosis, `CS Stage`) %>%
  mutate(se = sqrt(n))

# Step 2: Plot with error bars
ggplot(stage_summary, aes(x = `CS Stage`, y = n, fill = Diagnosis)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = n - se, ymax = n + se),
                position = position_dodge(width = 0.9),
                width = 0.3,
                color = "black") +
  labs(
    title = "Counts of CS Stage by Diagnosis + Standard Error",
    x = "CS Stage",
    y = "Number of Cases",
    fill = "Diagnosis"
  ) +
  scale_fill_manual(values = c(
    "CS" = "#800000",
    "MCS" = "#A9A9A9"
  )) +
  theme_minimal()
```
This is the table representation.
```{r}

stage_by_diagnosis <- Total_Data %>%
  filter(!is.na(`CS Stage`), !is.na(Diagnosis), Diagnosis != "Other") %>%  # 👈 excludes "Other"
  count(Diagnosis, `CS Stage`) %>%
  pivot_wider(names_from = `CS Stage`, values_from = n, values_fill = 0)


stage_table_percent <- stage_by_diagnosis %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  rowwise() %>%
  mutate(across(where(is.numeric), ~ paste0(.x, " (", round(100 * as.numeric(.x) / Total, 1), "%)"))) %>%
  ungroup()

chisq_matrix <- stage_by_diagnosis %>%
  column_to_rownames("Diagnosis") %>%
  as.matrix()

# Use Fisher's Exact Test
test_result <- fisher.test(chisq_matrix)
p_value <- format.pval(test_result$p.value, digits = 3, eps = 0.001)

library(knitr)
library(kableExtra)

caption_text <- paste("Distribution of CS Stage by Diagnosis (Excluding 'Other') — Fisher's Exact Test p =", p_value)


kable(stage_table_percent, caption = caption_text) %>%
  kable_styling()

stage_summary %>% arrange(desc(se)) %>% head()

```

# Analyze Survival and Expiration by CS Stage

```{r}
library(dplyr)
library(tidyr)
library(knitr)

# Step 1: Filter to only CS Stage D and E, count, and compute percentages
survival_by_cs_stage_table <- Total_Data %>%
  filter(CS_Stage %in% c("D", "E"), !is.na(`Pt outcome`)) %>%
  count(CS_Stage, `Pt outcome`) %>%
  group_by(CS_Stage) %>%
  mutate(percent = round(100 * n / sum(n), 1),
         Label = paste0(n, " (", percent, "%)")) %>%
  select(CS_Stage, `Pt outcome`, Label) %>%
  pivot_wider(names_from = `Pt outcome`, values_from = Label, values_fill = "0 (0%)")

# Step 2: Display table with counts and percentages
kable(survival_by_cs_stage_table, caption = "Patient Outcome by CS Stage (D and E Only)")

# Step 3: Prepare raw counts for test
survival_matrix <- Total_Data %>%
  filter(CS_Stage %in% c("D", "E"), !is.na(`Pt outcome`)) %>%
  count(CS_Stage, `Pt outcome`) %>%
  pivot_wider(names_from = `Pt outcome`, values_from = n, values_fill = 0) %>%
  column_to_rownames("CS_Stage") %>%
  as.matrix()

# Step 4: Run test
  test_result <- fisher.test(survival_matrix)
  test_name <- "Fisher's Exact Test"


# Step 5: Print p-value
cat(paste0("\n\n**", test_name, " p-value: ", format(test_result$p.value, digits = 3), "**"))

```


```{r}
library(ggplot2)
library(dplyr)

# Step 1: Prepare data and calculate standard error
plot_data <- Total_Data %>%
  filter(CS_Stage %in% c("D", "E"), !is.na(`Pt outcome`)) %>%
  count(CS_Stage, `Pt outcome`) %>%
  group_by(CS_Stage) %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    proportion = n / sum(n),
    se = sqrt(proportion * (1 - proportion) / sum(n)) * 100
  ) %>%
  ungroup()

# Step 2: Create the bar plot with error bars
ggplot(plot_data, aes(x = CS_Stage, y = percent, fill = `Pt outcome`)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = percent - se, ymax = percent + se),
    width = 0.3,
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  labs(
    title = "Patient Outcomes by CS Stage (D & E Only) + Standard Error",
    x = "CS Stage",
    y = "Percentage of Patients",
    fill = "Outcome"
  ) +
  scale_fill_manual(values = c(
    "expired" = "#800000",
    "survived" = "#A9A9A9"
  )) +
  theme_minimal()


```







#Table representation of survival and expired by diagnosis
Make a table with survival and expired on the y and then CS and MCS on the x
```{r}
# Your original full matrix
fisher_matrix <- matrix(c(37, 24, 37, 12, 41, 23), ncol = 2, byrow = TRUE)
rownames(fisher_matrix) <- c("CS", "MCS", "Other")
colnames(fisher_matrix) <- c("expired", "survived")

# Double-check structure
print(fisher_matrix)

# CS vs MCS
cs_mcs_matrix <- fisher_matrix[c("CS", "MCS"), , drop = FALSE]
fisher_cs_mcs <- fisher.test(cs_mcs_matrix)
cat("Fisher's Exact Test (CS vs MCS): p =", round(fisher_cs_mcs$p.value, 3), "\n")

# CS vs Other
cs_other_matrix <- fisher_matrix[c("CS", "Other"), , drop = FALSE]
fisher_cs_other <- fisher.test(cs_other_matrix)
cat("Fisher's Exact Test (CS vs Other): p =", round(fisher_cs_other$p.value, 3), "\n")

# MCS vs Other
mcs_other_matrix <- fisher_matrix[c("MCS", "Other"), , drop = FALSE]
fisher_mcs_other <- fisher.test(mcs_other_matrix)
cat("Fisher's Exact Test (MCS vs Other): p =", round(fisher_mcs_other$p.value, 3), "\n")
```
```{r}
# Step 1: Count and calculate percentages
mortality_diagnosis_table <- Total_Data %>%
  filter(!is.na(`Pt outcome`), !is.na(Diagnosis)) %>%
  count(Diagnosis, `Pt outcome`) %>%
  group_by(Diagnosis) %>%
  mutate(Percent = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  mutate(Label = paste0(n, " (", Percent, "%)")) %>%
  select(Diagnosis, `Pt outcome`, Label) %>%
  pivot_wider(names_from = `Pt outcome`, values_from = Label, values_fill = "0 (0%)")

# Step 2: Display formatted table with percentages
kable(mortality_diagnosis_table, caption = "Patient Outcomes by Diagnosis (Count and %)")

# Step 3: Create matrix with raw counts for Fisher's test
fisher_matrix <- Total_Data %>%
  filter(!is.na(`Pt outcome`), !is.na(Diagnosis)) %>%
  count(Diagnosis, `Pt outcome`) %>%
  pivot_wider(names_from = `Pt outcome`, values_from = n, values_fill = 0) %>%
  column_to_rownames("Diagnosis") %>%
  as.matrix()

# Step 4: Perform Fisher's Exact Test
fisher_result <- fisher.test(fisher_matrix)

# Step 5: Print test result
print(fisher_result)
```

Graph representation of survival and expired by diagnosis

```{r}
library(dplyr)
library(ggplot2)

# Step 1: Prepare data with standard error
plot_data <- Total_Data %>%
  filter(!is.na(`Pt outcome`), !is.na(Diagnosis)) %>%
  count(Diagnosis, `Pt outcome`) %>%
  group_by(Diagnosis) %>%
  mutate(
    Total = sum(n),
    Proportion = n / Total,
    Percent = 100 * Proportion,
    SE = sqrt(Proportion * (1 - Proportion) / Total) * 100,
    ymin = Percent - SE,
    ymax = Percent + SE
  ) %>%
  ungroup()

# Step 2: Plot with error bars
ggplot(plot_data, aes(x = Diagnosis, y = Percent, fill = `Pt outcome`)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax),
                position = position_dodge(width = 0.9),
                width = 0.2,
                color = "black") +
  labs(
    title = "Patient Outcomes by Diagnosis + Standard Error",
    x = "Diagnosis",
    y = "Percentage of Patients",
    fill = "Outcome"
  ) +
  scale_fill_manual(values = c(
    "expired" = "#800000",
    "survived" = "#A9A9A9"
  )) +
  theme_minimal()

```


# Tablle representation of Treatment by Diagnosis
```{r}
# List of treatment columns
treatment_vars <- c("ECMO", "Impella", "Coronary angiography or PCI",
  "RHC/hemodynamics", "TandemHeart",
  "RVAD", "IABP", "Surgical LVAD", "Open heart"
)
# Convert "Y" to 1, "N" to 0
Total_Data <- Total_Data %>%
  mutate(across(all_of(treatment_vars), ~ case_when(
    . == "Y" ~ 1,
    . == "N" ~ 0,
    TRUE ~ NA_real_
  )))

# Get all unique diagnoses
diagnoses <- unique(na.omit(Total_Data$Diagnosis))

# Create summary table
summary_table <- data.frame()

for (treatment_col in treatment_vars) {
  # Filter and count
  counts <- Total_Data %>%
    filter(!is.na(Diagnosis), !is.na(.data[[treatment_col]])) %>%
    group_by(Diagnosis) %>%
    summarise(
      n_total = n(),
      n_yes = sum(.data[[treatment_col]] == 1),
      .groups = "drop"
    ) %>%
    mutate(percent = round(100 * n_yes / n_total, 1),
           label = paste0(n_yes, " (", percent, "%)")) %>%
    select(Diagnosis, label)
  
  # Reshape to wide format
  row <- pivot_wider(counts, names_from = Diagnosis, values_from = label)
  row$Treatment <- treatment_col
  
  # Prepare matrix for statistical test
  test_matrix <- Total_Data %>%
    filter(!is.na(Diagnosis), !is.na(.data[[treatment_col]])) %>%
    count(Diagnosis, .data[[treatment_col]]) %>%
    pivot_wider(names_from = all_of(treatment_col), values_from = n, values_fill = 0) %>%
    column_to_rownames("Diagnosis") %>%
    as.matrix()
  
  # Run Fisher's Exact Test
  if (ncol(test_matrix) == 2 && all(rowSums(test_matrix) > 0)) {
    p_val <- round(fisher.test(test_matrix)$p.value, 3)
  } else {
    p_val <- NA
  }
  
  row$p_value <- p_val
  
  # Append row to summary table
  summary_table <- bind_rows(summary_table, row)
}

# Reorder columns: Treatment first, then diagnoses, then p-value
diagnosis_cols <- diagnoses[diagnoses %in% colnames(summary_table)]
summary_table <- summary_table %>%
  select(Treatment, all_of(diagnosis_cols), p_value)

# Display as table
kable(summary_table, caption = "Treatment Usage by Diagnosis with Fisher's Exact Test p-values") %>%
  kable_styling(full_width = FALSE)

```

Graph representation of Treatment by Diagnosis
```{r}

# Step 1: Extract percent values from summary_table (remove text like "12 (25%)")
percent_data <- summary_table %>%
  select(-p_value) %>%
  pivot_longer(cols = -Treatment, names_to = "Diagnosis", values_to = "Label") %>%
  mutate(
    Percent = as.numeric(str_extract(Label, "(?<=\\().*?(?=%)"))  # Extract number inside the parentheses
  )

# Step 2: Plot grouped bar chart
ggplot(percent_data, aes(x = Treatment, y = Percent, fill = Diagnosis)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Percent of Patients Receiving Each Treatment by Diagnosis",
    x = "Treatment",
    y = "Percent Receiving Treatment",
    fill = "Diagnosis"
  ) +
  scale_fill_manual(values = c(
    "CS" = "#800000",
    "MCS" = "#A9A9A9",
    "Other" = "#B46A55"
  )) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```



# Calculate Average Support per Diagnosis Type
```{r}
# Define relevant treatment columns to include in the support score
treatment_cols <- c(
  "ECMO", "Impella", "Coronary angiography or PCI", "RHC/hemodynamics",
  "Open heart", "TandemHeart", "RVAD", "IABP", "Surgical LVAD", "transplant"
)

# Calculate support score for each patient
# Convert 'Y' to 1 and others to 0 for selected treatment columns
support_data_diagnosis <- Total_Data %>%
  rowwise() %>%
  mutate(support_score = sum(c_across(all_of(treatment_cols)) == "Y", na.rm = TRUE)) %>%
  ungroup()

# Calculate average support score and SE per Diagnosis Type
avg_support_by_diagnosis <- support_data_diagnosis %>%
  filter(!is.na(Diagnosis)) %>%
  group_by(Diagnosis) %>%
  summarise(
    Average_Support = mean(support_score, na.rm = TRUE),
    se = sd(support_score, na.rm = TRUE) / sqrt(n())
  ) %>%
  arrange(Diagnosis)

# Display the summary table
knitr::kable(avg_support_by_diagnosis, caption = "Average Support Score by Diagnosis Type")

# Perform ANOVA test
anova_result_diagnosis <- aov(support_score ~ Diagnosis, data = support_data_diagnosis %>% filter(!is.na(Diagnosis)))

# Print ANOVA summary
cat("\n### ANOVA Test Results:\n")
print(summary(anova_result_diagnosis))
```

#Graph
```{r}
# Create the bar plot
support_plot_diagnosis <- ggplot(avg_support_by_diagnosis, aes(x = Diagnosis, y = Average_Support, fill = Diagnosis)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = Average_Support - se, ymax = Average_Support + se), width = 0.2, position = position_dodge(0.9)) +
  labs(
    title = "Average Number of Treatments per Patient by Diagnosis Type + Standard Error",
    x = "Diagnosis Type",
    y = "Average Number of Treatments"
  ) +
  scale_fill_manual(values = c(
    "CS" = "#800000",
    "MCS" = "#A9A9A9",
    "Other" = "#D9D9D9"
  )) +
  theme_minimal() +
  theme(legend.position = "none")

# Print the plot to the HTML report
print(support_plot_diagnosis)

# Save the plot to a file
ggsave(here("results", "average_support_by_diagnosis.png"), plot = support_plot_diagnosis, width = 8, height = 6, dpi = 300)

```



```{r}
# Define treatment columns
treatment_cols <- c(
  "ECMO", "Impella", "Coronary angiography or PCI", "RHC/hemodynamics",
  "Open heart", "TandemHeart", "RVAD", "IABP", "Surgical LVAD", "transplant"
)

# Step 1: Create per-patient support score
support_data_stage <- Total_Data %>%
  rowwise() %>%
  mutate(support_score = sum(c_across(all_of(treatment_cols)) == "Y", na.rm = TRUE)) %>%
  ungroup() %>%
  filter(CS_Stage %in% c("D", "E"))  # Keep only CS stages D and E

# Step 2: Summary table with mean and SE
support_data_stage %>%
  group_by(CS_Stage) %>%
  summarise(
    `Average Number of Treatments` = mean(support_score, na.rm = TRUE),
    `Standard Error (SE)` = sd(support_score, na.rm = TRUE) / sqrt(n())
  ) %>%
  rename(`CS Stage` = CS_Stage) %>%
  knitr::kable(
    caption = "Table: Average Number of Treatments per Patient with Standard Error for CS Stages D and E",
    digits = 2
  )

# Step 3: Run ANOVA
anova_result <- aov(support_score ~ CS_Stage, data = support_data_stage)

# Step 4: Print p-value
cat("\n### ANOVA Test Results (CS Stage D vs E):\n")
print(summary(anova_result))
```


```{r}
# Calculate Average Support per CS Stage
# Filter only stages C and D
filtered_data <- avg_support_by_cs_stage %>%
  filter(CS_Stage %in% c("E", "D"))

# Create the bar plot
support_plot <- ggplot(filtered_data, aes(x = CS_Stage, y = Average_Support, fill = CS_Stage)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = Average_Support - se, ymax = Average_Support + se), 
                width = 0.2, position = position_dodge(0.9)) +
  labs(
    title = "Average Number of Treatments per Patient (CS Stage D & E)",
    x = "CS Stage",
    y = "Average Number of Treatments"
  ) +
   scale_fill_manual(values = c(
    "E" = "#800000",
    "D" = "#A9A9A9"
  )) +
  theme_minimal() +
  theme(legend.position = "none")

# Print the plot
print(support_plot)

# Save the plot
ggsave(here("results", "average_support_by_cs_stage_C_D.png"), 
       plot = support_plot, width = 8, height = 6, dpi = 300)

```









