---
title: "Data analysis and model fitting for PSOMER data"
output: html_notebook
---

```{r}
# for data manipulation and visualization
library(tidyverse)

# for flexible paths and reproducible research
library(here)
```

## Load in dataset

```{r}
data <- read_csv(here("data", "final_cohort.csv"))
```

## Data exploration

```{r}
data
```

## Data exploration figures

## Model fitting

```{r}
data %>%
  #break age into quintiles
  mutate(age_decile = cut(age, 10)) %>%
  group_by(age_decile) %>%
  summarise(outcome_rate = mean(outcome)) %>% 
  ggplot(aes(x= age_decile, y = outcome_rate)) + geom_point()

```

## Hypothesis

**Age is an independent risk factor for myocardial infarction.**

## Statistical model

$$log(\frac{P(Y =1)}{(1- P(Y=1)}) = \beta_0 + \beta_1*age + \beta_2*diabetes$$

### Null hypothesis

**There is NO association between age and myocardial infarction**

$$\beta_1 = 0$$

```{r}
glm(outcome ~ age + diabetes, data = data, family = "binomial") %>% summary()
```

The coefficient for age is significant at p \< 0.05, indicating we can *reject* the null hypothesis in favor of the alternative hypothesis

# Interpretation of effect size using inline r code
```{r}
age_coefficient <- glm(outcome ~ age + diabetes, data = data, family = "binomial") %>% 
  summary() %>% 
  .$coefficients %>% 
  .[2,1]

exp(age_coefficient)
```

The odds ratio for age is `r round(exp(age_coefficient), digits =2)`, indicating that for each additional year of age, the odds of having a myocardial infarction increase by `r round(100*(exp(age_coefficient)-1))`%.